Legion: Expressing Locality and Independence with Logical Regions

### 2012-04-27 doc/sc2012/legion.tex
secs 1 and 2

\begin{abstract}
Modern parallel architectures have both heterogeneous processors and
deep, complex memory hierarchies.  We present Legion, a programming
model and runtime system for achieving high performance on these
machines.  Legion is organized around {\em logical regions}, which
express both locality and independence of program data and {\em
  tasks}, functions that perform computations on regions.  We describe
a runtime system that dynamically extracts parallelism from
Legion programs, using a distributed, parallel scheduling algorithm
that identifies both independent tasks and nested parallelism.  Legion
also enables explicit, programmer controlled movement of data through
the memory hierarchy and placement of tasks based on locality
information via a novel mapping interface.  We evaluate our Legion
implementation using four applications: fluid-flow on a regular grid,
a three-level AMR code solving a simple heat transfer equation, a circuit simulation,
and LINPACK.
%Running on a 4 node cluster with 8 total GPUs 
%and 4 levels of memory hierarchy, our implementation of Legion 
%achieves up to 5.9X speedup over a single CPU-GPU node on real-world applications.
\end{abstract}

@@

In this paper we describe Legion, a parallel programming system based
on using {\em logical regions} to describe the organization of data.
A logical region names a set of objects.  Logical regions are first
class values in Legion, and in particular may be passed as arguments
to distinguished functions called {\em tasks} that access the data in those regions, providing locality
information.  Logical regions may be {\em partitioned} into (usually)
disjoint {\em subregions}, providing information for determining independence of computations.  Furthermore,
computations access logical regions with particular {\em privileges} (e.g., {\em
  read-only} and {\em read-write}, among others) 
and {\em coherence} (e.g., {\em exclusive access} and {\em atomic access}, among others).
Priveleges express how a task may use its region arguments, providing data dependence information
that is used to guide the extraction of parallelism (e.g., if two tasks access the same region with
{\em read-only} priveleges the two tasks can potentially be run in parallel), and coherence annotations
express the required semantics of concurrent accesses to the region.  For example, if tasks $x$ and $y$ access region $r$
with {\em exclusive} access, then the effects of $x$ and $y$ on $r$ are guaranteed to be the same as if they had executed
in the sequential order expressed by the program; on the other hand, if the tasks access $r$ with {\em atomic} coherence, then 
Legion guarantees only  atomicity of the tasks with respect to $r$: either $x$ appears to run entirely before $y$ or vice versa.

To introduce the programming model, we begin with an extended example of a circuit simulation programmed in Legion
in Section~\ref{sec:ex}. This example illustrates regions, tasks, permissions/coherence properties, the interactions
between them, and how these building blocks are assembled into a realistic Legion program.   Subsequent sections each
describe a contribution in the implementation and evaluation of Legion:

@@

\item We define a {\em software out-of order processor}, or SOOP, for scheduling tasks with region arguments in a manner
analogous to how out-of-order hardware schedulers process instructions with register arguments.  In addition to pipelining
the scheduling of tasks over several stages, our SOOP is also hierarchical and can extract nested parallelism (because tasks may recursively spawn subtasks) as well as concurrent and distributed across the machine (Section~\ref{sec:soop}).

\item A crucial part of orchestrating a program's execution is how
tasks are assigned (or {\em mapped}) to processors and how regions are
mapped to specific memory units within the machine.  Often there is
application-specific information that, if exploited, results in better
mappings than a generic, one-size-fits-all mapping strategy.  We describe
a {\em mapping interface} that allows programmers to give the SOOP a specification
of how to map tasks and regions.  This mapping API is carefully defined so that
any use-defined mapping strategy can only affect the performance of applications,
not their correctness.  Thus, user mappings only affect the time and resources required
to run the program and cannot change the program's output (Section~\ref{sec:mapping}).

\item We present the results of experiments on four applications: fluid-flow on a regular grid,
a three-level AMR code solving a simple heat transfer equation, a circuit simulation,
and LINPACK.  We run each of these four codes and best available reference versions on three different
clusters of multicore processors with attached GPUs: a small cluster of our own, a medium-sized cluster designed for visualization applications, and up to half of the Keeneland supercomputer (Section~\ref{sec:exp}).

\end{itemize}
\input{code_ex}
\input{circuit}
\input{part_fig}
\input{soop}
\input{mapping}cir
\input{experiments}
\input{related}
\input{conclusion}

>>>>

\begin{abstract}
Modern parallel architectures have both heterogeneous processors and
deep, complex memory hierarchies.  We present Legion, a programming
model and runtime system for achieving high performance on these
machines.  Legion is organized around {\em logical regions}, which
express both locality and independence of program data, and {\em
  tasks}, functions that perform computations on regions.  We describe
a runtime system that dynamically extracts parallelism from
Legion programs, using a distributed, parallel scheduling algorithm
that identifies both independent tasks and nested parallelism.  Legion
also enables explicit, programmer controlled movement of data through
the memory hierarchy and placement of tasks based on locality
information via a novel mapping interface.  We evaluate our Legion
implementation using four applications: fluid-flow on a regular grid,
a three-level AMR code solving a standard heat diffusion equation, a circuit simulation,
and LINPACK.
%Running on a 4 node cluster with 8 total GPUs 
%and 4 levels of memory hierarchy, our implementation of Legion 
%achieves up to 5.9X speedup over a single CPU-GPU node on real-world applications.
\end{abstract}

@@

In this paper we describe Legion, a parallel programming system based
on using {\em logical regions} to describe the organization of data.
A logical region names a set of objects.  Logical regions are first
class values in Legion, and in particular may be passed as arguments
to distinguished functions called {\em tasks} that access the data in
those regions, providing locality information.  Logical regions may be
{\em partitioned} into (usually) disjoint {\em subregions}, providing
information for determining independence of computations.
Furthermore, computations access logical regions with particular {\em
  privileges} (e.g., {\em read-only} and {\em read-write}, among
others) and {\em coherence} (e.g., {\em exclusive access} and {\em
  atomic access}, among others).  Priveleges express how a task may
use its region arguments, providing data dependence information that
is used to guide the extraction of parallelism (e.g., if two tasks
access the region $r$with {\em read-only} priveleges the two tasks
can potentially be run in parallel, at least with respect to $r$),
and coherence properties express the required semantics of concurrent
accesses to the region.  For example, if tasks $x$ and $y$ access
region $r$ with {\em exclusive} access, then the effects of $x$ and
$y$ on $r$ are guaranteed to be the same as if they had executed in
the sequential order expressed by the program; on the other hand, if
the tasks access $r$ with {\em atomic} coherence, then Legion
guarantees only atomicity of the tasks with respect to $r$: either $x$
appears to run entirely before $y$ or vice versa.

To introduce the programming model, we begin with an extended example
of a circuit simulation programmed in Legion in
Section~\ref{sec:ex}. This example illustrates regions, tasks,
permissions and coherence properties, the interactions between them,
and how these building blocks are assembled into a Legion
program.  Subsequent sections each describe a contribution in the
implementation and evaluation of Legion:

@@

\item We define a {\em software out-of-order processor}, or SOOP, for
  scheduling tasks with region arguments in a manner analogous to how
  out-of-order hardware schedulers process instructions with register
  arguments.  In addition to pipelining the execution of tasks over
  several stages, our SOOP is hierarchical and naturally extracts nested
  parallelism (because tasks may recursively spawn subtasks) as well
  as concurrent and distributed across the machine
  (Section~\ref{sec:soop}).

\item A crucial part of orchestrating a program's execution is how
tasks are assigned (or {\em mapped}) to processors and how regions are
mapped to specific memory units within the machine.  Often there is
application-specific information that, if exploited, results in better
mappings than a generic, one-size-fits-all mapping strategy.  We describe
a {\em mapping interface} that allows programmers to give the SOOP a specification
of how to map tasks and regions for a specific application, or even part of an
application.  This mapping API is carefully defined so that
any user-supplied mapping strategy can only affect the performance of applications,
not their correctness.  Thus, user mappings only affect the time and resources required
to run the program and cannot change the program's output (Section~\ref{sec:mapping}).

\item We present the results of experiments on four applications:
  fluid-flow on a regular grid, a three-level AMR code solving a
  standard heat diffusion equation, a circuit simulation, and LINPACK.
  We run each of these four codes and best available reference
  versions on three different clusters of multicore processors with
  attached GPUs: a small cluster of our own (4 nodes with 2 6-core CPUs and 2 GPUs per node), a medium-sized cluster
  designed for visualization applications (10 nodes with 2 6-core CPUs and 5 GPUs per node), and up to a third of the
  Keeneland supercomputer (32 nodes, 2 6-core CPUs and 3 GPUs per node, see Section~\ref{sec:exp}).

\end{itemize}
\input{code_ex}
\input{circuit}
\input{part_fig}
\input{soop}
\input{mapping}
\input{experiments}
\input{related}
\input{conclusion}
